---
title: "Problem Set 1"
author: "Matt Ostendorf"
subtitle: Comp 4441
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
library(HistData)
library(tidyverse)

```

# Introduction

These questions were rendered in R markdown through RStudio (<https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>, <http://rmarkdown.rstudio.com> ).

Please generate your solutions in R markdown. From an Rmd file in RStudio, you can generate a word document by selecting the "Knit to Word" option next to the "Knit" icon in the toolbar above the edit window. Please upload both a knitted doc, docx, or pdf document in addition to the Rmd file.

Please put your name in the "author" section in the header.

RStudio may ask you to install packages when you run this code. Accepting the request will allow you to proceed.

In this problem set, most of the code is provided. The challenge is to interpret the results according to the principles introduced in the polio case study in week 1.

# Load Data

```{r}
data("PolioTrials")
dat<-PolioTrials
kable(dat[,1:4])

```


# Question 1

Please carry out the analysis below and answer the questions that follow. 

## Context

The basic question "did the vaccine work?" was addressed in week 1 using the data from the randomized control trial of the Salk vaccine. The count of paralytic polio cases in the vaccinated group was compared to the counts that were produced under the null hypothesis that the vaccine had no effect on the incidences of paralytic polio. The second model examined in week 1 for this null hypothesis (01_polio_simulation_binomial_model.Rmd) was that each paralytic polio case in the pooled treatment and placebo group was assigned to the treatment group with probability equal to the ratio of the size of the treatment group to the size of pooled treatment and placebo group.   

Below, the computation used in that analysis is repeated for the vaccinated group and the control group in the observed control trial. (`Vaccinated` and `Controls` in the `ObservedControl` experiment)

Recall rbinom(n,ct,prop) is a function that models the number of random assignments to the distinguished group from a population of size `ct` if the probability of assignment to the distinguished group is `prop`. The value of `n` is the number of times to repeat the experiment.

### Test

```{r}
n<-10000 # number of simulations

# Calculate the number of paralytic polio cases in the pooled vaccination and control group.
ct<-sum(dat$Paralytic[5:6])

# Calculate the proportion "prop" of the the pooled vaccination and control group that are in the vaccination group.
prop<-dat$Population[5]/sum(dat$Population[5:6])

# Generate 10,000 counts of paralytic polio cases in the vaccination group under the model that each paralytic polio case in the pooled vaccination and control group has probability "prop" of being assigned to the vaccination group.
set.seed(45678765)
sim<-rbinom(n,ct,prop)

# Plot a histogram of the simulated counts.
qplot(sim,bins=20)

# Calculate the proportion of the simulated counts of paralytic polio in the "Vaccinated" group that are less than or equal to the observed count:
mean(sim<=dat$Paralytic[5])
```

## Q1, part 1
(10 points)

*  What is the proportion of paralytic polio cases in the `Vaccinated` group in the `ObservedControl` experiment? 
0.0001711727

*  What is the proportion of paralytic polio cases in the `Controls` group in the `ObservedControl` experiment? 
0.0004550638

*  What is the proportion of paralytic polio cases in the pooled `Vaccinated` and `Controls` groups in the `ObservedControl` experiment? 
0.0003885254

The following computations may be helpful.

```{r}
# ratio of the value in the 5th row of the "Paralytic"
# column of "dat" to the value in the 5th row of the "Population"
# column of "dat":
str_glue("The proportion of paralytic polio cases in the `Vaccinated` group in the `ObservedControl` experiment is {dat$Paralytic[5]/dat$Population[5]}")

# ratio of the value in the 6th row of the "Paralytic"
# column of "dat" to the value in the 6th row of the "Population"
# column of "dat":
str_glue("The proportion of paralytic polio cases in the `Controls` group in the `ObservedControl` experiment is {dat$Paralytic[6]/dat$Population[6]}")

# ratio of the sum of the values in the 5th and 6th row of 
# the "Paralytic" column of "dat" to the sum of the values in the 
# 5th and 6th row of the "Population" column of "dat":
str_glue("The proportion of paralytic polio cases in the pooled `Vaccinated` and `Controls` groups in the `ObservedControl` experimentsum is {sum(dat$Paralytic[5:6])/sum(dat$Population[5:6])}")
```

## Q1, part 2
(15 points)

Is the observed number of paralytic polio cases in the `Vaccinated` group in the `ObservedControl` experiment consistent with the probability model that each paralytic polio case in the pooled vaccinated and control group was assigned to the vaccinated group with probability equal to the ratio of the size of the vaccinated group to the size of pooled vaccinated and control group? 

```{r}
# some statistics of the simulated values under the probability model
mean(sim)
min(sim)
max(sim)
```

No it is not consistent. The `Vaccinated` group in the `ObservedControl` group had 38 cases of paralytic polio. When we run the probability model 10,000 times we are seeing a mean of 86 and a min of 54. The min of 54 in the model is greater than the 38 observed cases. This provides further evidence that it is not consistent.


## Q1, part 3
(10 points)

Using your conclusion in part 2, can the data from the `ObservedControl` experiment be interpreted as evidence that the vaccination *causes* a reduction in the likelihood of contracting paralytic polio? Please explain. Recall that the `Vaccinated` group consists of second graders whose parents consented to vaccination while the `Controls` group consists of first and third graders.

We can not make the statement that the vaccine caused a reduction in the likelihood of contracting paralytic polio. This is because we have not controlled for the difference in student ages (i.e. what grade the student is in). We also haven't controlled for the parents decision to allow their child to be vaccinated. 


# Question 2

Please carry out the analysis below and answer the question that follows.

## Context

In this section, you will address the question of whether the `NotInoculated` and `Placebo` groups in the `Randomized Control` experiment had statistically significantly different rates of paralytic polio. 

Recall that the `NotInoculated` and `Placebo` groups differ in that the children in the `Placebo` group had been enrolled in the vaccine trial while the parents of the children in the `NotInoculated` group did not enroll their children. If the home environments in these groups differed systematically in ways related to susceptibility to paralytic polio, rates in the two groups could differ beyond the amount readily explained by chance. If the enrolled and not enrolled populations didn't differ in relevant ways, the differences should be fairly consistent with chance.

The approach, using the `rbinom` function again, implements the idea that populations in the `NotInoculated` and `Placebo` groups in the `Randomized Control` experiment were the same in regards to paralytic polio cases by using the `rbinom` function to assign paralytic polio cases in the combined `NotInoculated` and `Placebo` groups of the `RandomizedControl` experiment to the `Placebo` group with probability equal to the ratio of the size of the `Placebo` group to the size of pooled `Placebo` group and `NotInoculated` group.

### Data summary

```{r}
# proportion of paralytic polio cases in "Placebo"
str_glue("The proportion of paralytic polio cases in Placebo {dat$Paralytic[2]/dat$Population[2]}")

# proportion of paralytic polio cases in "NotInoculated"
str_glue("The proportion of paralytic polio cases in NotInoculated {dat$Paralytic[3]/dat$Population[3]}")

# proportion of paralytic polio cases in "Placebo" and "NotInoculated" combined
str_glue("The proportion of paralytic polio cases in placebo and NotInoculated combined {sum(dat$Paralytic[2:3])/sum(dat$Population[2:3])}")
```


### Test

```{r}
n<-10000 # number of simulations

# Calculate the number of paralytic polio cases in the pooled "Placebo" and "NotInoculated" group.
ct<-sum(dat$Paralytic[2:3])

# Calculate the proportion "prop" of the the pooled "Placebo" and "NotInoculated" group that are in the "Placebo" group.
prop<-dat$Population[2]/sum(dat$Population[2:3])

# Generate 10,000 counts of paralytic polio cases in the "Placebo" group under the model that each paralytic polio case in the pooled pooled "Placebo" and "NotInoculated" group has probability "prop" of being assigned to the "Placebo" group.
set.seed(45678765)
sim2<-rbinom(n,ct,prop) 
qplot(sim2,bins=20)

# proportion of the simulated counts of paralytic polio in the "Placebo" that are less than or equal to the observed count:
mean(sim2<=dat$Paralytic[2])

# proportion of the simulated counts of paralytic polio in the "Placebo" that are greater than or equal to the observed count:
mean(sim2>=dat$Paralytic[2])
```

# Q2
(15 points)

Consider the null hypothesis that children who receive the placebo and children whose parents don't consent to participation have the same susceptibility to paralytic polio. One possible probability model for this is that each case in the pooled `Placebo` and `NotInoculated` paralytic polio cases can be viewed as being assigned to the `Placebo` group with probability equal to the proportion of the `Placebo` population in the pooled `Placebo` and `NotInoculated` population. Are these data consistent with the null hypothesis? Please explain your conclusion and support it by referring to the values generated in simulation above. 

```{r}
# some statistics of the simulated values under the probability model
mean(sim2)
min(sim2)
max(sim2)
```

No they are not consistent with the null hypothesis. When we look at the randomized control placebo group we get 115 cases of paralytic polio and when we run the simulation we get a mean of 88. The 115 is well outside the mean of the simulation. We do see that 115 is in the range of the max of 118, but that is more of an outlier and should be considered random chance to archive than actually statistically significant. 

















